---
title: "Fecundity-at-age"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: FALSE
    code_folding: show
    number_sections: TRUE
---

# SET-UP

```{r markdown, echo=F}
library(knitr)
opts_chunk$set(echo = T, collapse = T, fig.align = "center", fig.width = 9, fig.height = 6)
options(width = 300)
```    

```{r settings, message = F}
source('0.0_settings.R')

this.year <- 2022
```

# Input data
## Read
```{r read, message = F}
# Pelletier data
pel <- read.csv('data/pelletier.csv')

# bio data
new <- FALSE
if(new){
    bio <- get.bio(species='maquereau',user=imlp.user,password=imlp.pass)
    save(bio,file=paste0('Rdata/bio_',Sys.Date(),'.Rdata'))
}else{
    df <- file.info(list.files("Rdata/", full.names = T,pattern="bio_"))
    load(rownames(df)[which.max(df$mtime)])
}
```

## Transformations
```{r transformations, message = F}
## Pelletier data
# add variables
pel$gsi <- with(pel,weight_gonads/weight*100)
pel$ageplus <- pmin(pel$age,10)

# melt dataframe
pelm <- pel[,c('id','fecundity','length','weight','age','weight_gonads','doy','gsi')]
pelm <- melt(pelm,id=c('id','fecundity'))

## bio data
# subset
bio$doy <- lubridate::yday(as.Date(paste(bio$year,bio$month,bio$day,sep='/')))
bio.faa <- bio[bio$matur==5 & 
               bio$sex=='F' & 
               bio$agef>0,
               c('year','month','matur','sex','wgonad','agef','weight','length.frozen','doy')]
bio.faa <- bio.faa[!is.na(bio.faa$agef) & !is.na(bio.faa$wgonad) & !is.na(bio.faa$weight),]
names(bio.faa)[c(6,8)] <- c('age','length')  # length kept to help identify potentially wrong ages

# transform/add
bio.faa$wgonad <- bio.faa$wgonad*1000 # same unit as pelletier (g)
bio.faa$weight <- bio.faa$weight*1000
bio.faa$gsi <- with(bio.faa,wgonad/weight*100)
bio.faa$ageplus <- pmin(bio.faa$age,10)
```

# Pelletier data

## Plots {.tabset}

### relationships
```{r plot_pel}
ggplot(pelm,aes(y=fecundity,x=value))+
    geom_point()+
    facet_wrap(~variable,scale='free_x')
```

### relationships (log)
```{r plot_pel_log}
ggplot(pelm,aes(y=fecundity,x=log(value)))+
    geom_point()+
    facet_wrap(~variable,scale='free_x')
```

### pairs
```{r plot_pelpairs,fig.width=16,fig.height=16}
d <- pel[,c('length','weight','age','weight_gonads','doy','gsi','fecundity')]
suppressWarnings(chart.Correlation(d, histogram = TRUE, method = "pearson"))
```

### pairs (log)
```{r plot_pelpairs_log,fig.width=16,fig.height=16}
suppressWarnings(chart.Correlation(log(d), histogram = TRUE, method = "pearson"))
```

## Linear models
Replicate Pelletier analyses. Pelletier found gonad weight to be the one with the highest predictive power.
```{r pel_mod}
# original analyses by Pelletier 1986
models <- ddply(pelm,c('variable'),function(x){
    m <- lm(log10(fecundity)~log10(value),data=x)
    c(variable=unique(as.character(x$variable)),
      intercept=round(unname(coef(m)[1]),2),
      slope=round(unname(coef(m)[2]),2),
      R2=round(summary(m)$r.squared,2),
      R2.adj=round(summary(m)$r.squared,2),
      sigma=round(summary(m)$sigma,2),
      n=nrow(x[!is.na(x$value),]))
}) 

kable(models[order(models$R2,decreasing = T),])
```

## Outlier removal
```{r pel_out}
# age
pelo <- pel[pel$age!=15 & !is.na(pel$age),]

# gonad weight
quantile(bio[bio$matur==5 & !is.na(bio$matur) & !is.na(bio$wgonad),'wgonad']*1000,c(0.005,0.995))
pelo <- pelo[pelo$weight_gonads<200,] # weights above 200g are unrealistic

nrow(pelo)
```

## New model

### Fit
Gonad weight and age are correlated, and therefore fecundity is predicted in function of age + GSI.
Mathieu Boudreau tested a range of models. See script in depricated folder. Here I only retained the selected one to use for model input. Selected robust regression because less weight to lower fecundity values (individuals for which spawning might have started already).

```{r pel_eq, message = F}
this.m <- robust::lmRob(log(fecundity) ~ log(gsi) + log(ageplus), data = pelo)

car::vif(this.m) # very low
```

### Model info {.tabset}

#### Summary
```{r pel_modsummary}
summary(this.m)
```

#### Validation 1
```{r plot_modvalid1}
par(mfrow=c(2,2))
plot(this.m,ask = FALSE)
```

#### Validation 2
```{r plot_mod_valid2}
p <-ggeffects::ggeffect(this.m)
grid.arrange(plot(p[[1]]),plot(p[[2]]),nrow=1)
```

# Bio data

## Outlier removal
```{r bio_out, message = F}
nrow(bio.faa)

# month (no "pre-spawners" after July)
bio.faa <- bio.faa[bio.faa$month <8,]

# remove fish with ages that cannot be trusted, based on their length (see 3.0_caa))
bio.faa <- bio.faa[bio.faa$age<18,]      
bio.faa <- ddply(bio.faa,c('age'),transform,outlierAL=outlier(length,coef=3))  
bio.faa[is.na(bio.faa$outlierAL),'outlierAL'] <- FALSE

# incorrect weights
bio.faa <- bio.faa[bio.faa$weight>0,]

lw <- lm(log(bio.faa$weight)~log(bio.faa$length))
bio.faa[!is.na(bio.faa$length),'outlierLW'] <- abs(residuals(lw))>sd(residuals(lw))*3
bio.faa[is.na(bio.faa$outlierLW),'outlierLW'] <- FALSE

# gsi
bio.faa <- bio.faa[bio.faa$gsi<33,] # one with gsi =35% (more than one third of this fish is gonad...)
bio.faa <- bio.faa[bio.faa$gsi>3.4,]  # smaller than ever observed for age 2. presumably wrong maturity stage
```

## Plots for outliers {.tabset}

### outliers AL
```{r plot out_al}
ggplot(bio.faa,aes(x=age,y=length,col=outlierAL))+geom_point()
bio.faa <- bio.faa[!bio.faa$outlierAL,]
bio.faa$outlierAL <- NULL
```

### outliers LW
```{r plot out_lw}
ggplot(bio.faa,aes(x=log(length),y=log(weight),col=outlierLW))+geom_point()
bio.faa <- bio.faa[!bio.faa$outlierLW,]
bio.faa$outlierLW <- NULL
```

### outliers GSI
```{r plot out_gsi}
ggplot(bio.faa,aes(x=as.factor(ageplus),y=gsi))+geom_boxplot()
```

## Comparison Pelletier
```{r data_comp}
names(bio.faa)[which(names(bio.faa)=='wgonad')] <- 'weight_gonads'
bio.faa$data <- 'bio'
pel$data <- 'pelletier'
cols <- c('weight','age','weight_gonads','gsi','data')
comb <- rbind(bio.faa[,cols],pel[,cols])
comb$id <- 1:nrow(comb)

combm <- melt(comb,id=c('id','data'))
ggplot(combm,aes(x=data,y=value,col=data))+geom_boxplot()+facet_wrap(~variable,scale='free_y')
```

## Which months to include?
There is a clear relationship between fecundity and doy.
Cannot add variable in model (correlated with gsi). 
If there are more samples from early on, might result in overestimate of fecundity. 

Kept july samples (no correct for doy either) because they are rare and have negligible impact.

### Plots   {.tabset}

#### gsi vs month
```{r pred_months}
n <- ddply(bio.faa,c('ageplus','month'),summarise,n=length(year))
ggplot(bio.faa,aes(x=as.factor(month),y=gsi))+
    geom_boxplot(aes(fill=as.factor(month)))+
    geom_text(data=n,aes(y=Inf,label=n),vjust=1.1)+
    facet_wrap(~ageplus)
```

#### sample date over time
```{r pred_year}
dates <- ddply(bio.faa,c('year'),summarise,mu=mean(doy),low=mu-2*sd(doy),high=mu+2*sd(doy))
ggplot(dates,aes(x=year,y=mu))+geom_point()+geom_errorbar(aes(ymin=low,ymax=high))+
    labs(y='Average sample date',x='Year')
```

# Predictions 

## Predict
See script test prediction to help understand the approach taken here. se is only used to help weigh the smoothing (but is relative because various shrinkage factors can be used).
```{r pred_comp}
faa <- ddply(bio.faa,c('year','ageplus'),summarise,n=length(gsi),gsi.sd=sd(gsi),gsi=mean(gsi))

faa$fec <- exp(predict(this.m,faa))
faa[!is.na(faa$gsi.sd),'fec.se'] <- exp(predict(this.m,data.frame(ageplus=faa$ageplus,gsi=faa$gsi.sd)))
faa[!is.na(faa$gsi.sd),'low'] <- exp(predict(this.m,data.frame(ageplus=faa$ageplus,gsi=pmax(faa$gsi-2*faa$gsi.sd,0.1))))
faa[!is.na(faa$gsi.sd),'high'] <- exp(predict(this.m,data.frame(ageplus=faa$ageplus,gsi=pmax(faa$gsi+2*faa$gsi.sd,0.1))))

minf <- 5 # number of fish should be higher than this
```

## Plots {.tabset}

### FAA raw
```{r plot_faaraw}
ggplot(faa[faa$n>=minf,],aes(x=year,y=fec))+
    geom_point()+
    geom_line()+
    facet_wrap(~ageplus)
```

### FAA vs pelletier
```{r plot_faarawpel}
pelfaa <- ddply(pelo[!is.na(pelo$ageplus),],c('ageplus'),summarise,fec=mean(fecundity),fec.se=sd(fecundity))
ggplot(faa[faa$n>=minf,])+
    geom_rect(data=pelfaa,aes(ymin=fec-2*fec.se,ymax=fec+2*fec.se,xmin=-Inf,xmax=Inf),fill='orange',alpha=0.5)+
    geom_hline(data=pelfaa,aes(yintercept=fec),col='orange')+
    geom_point(aes(x=year,y=fec))+
    geom_line(aes(x=year,y=fec))+
    facet_wrap(~ageplus)+
    labs(y='Fecundity',x='Year')
```

### predicted fec-gsi
```{r plot_predfecgsi}
ggplot(faa,aes(x=gsi,y=fec))+
    geom_point(data=pelo,aes(y=fecundity),col='orange')+
    geom_point()+
    labs(y='Fecundity',x='GSI')
```

### predicted fec-age
```{r plot_predfecage}
ggplot(faa,aes(x=as.factor(ageplus),y=fec))+
    geom_boxplot(data=pelo[!is.na(pelo$ageplus),],aes(y=fecundity,group=ageplus),fill='orange',col='darkgoldenrod')+
    geom_point()+
    labs(y='Fecundity',x='Age')
```

### n
```{r plot_n}
ggplot(faa,aes(x=year,y=as.factor(ageplus)))+
    geom_tile(aes(fill=n))+
    geom_text(aes(label = n), size = 2)+
    scale_fill_gradient(low = "orange", high = "darkred") + 
    labs(y='Age',x='Year')+
    scale_x_continuous(expand = c(0,0))+
    scale_y_discrete(expand=c(0,0))
```

### GSI
Check underlying trend
```{r plot_gsiyear}
ggplot(bio.faa,aes(x=year,y=gsi))+
    geom_boxplot(aes(group=year))+
    geom_smooth()+
    facet_wrap(~ageplus)+
    labs(x='Year',y='GSI (%)')+
    theme_minimal()
```


# Fill and smooth 

## fit model
```{r faa_fit, message = F, fig.height = 11,fig.width = 10,results = 'hide',message=FALSE}
faam <- faa[faa$n>=minf,]
faam$cv <- with(faam,fec.se/fec)
names(faam)[which(names(faam)=='ageplus')] <- 'age'

all <- expand.grid(year=1968:this.year, age=min(faam$age):max(faam$age))
faam <- merge(all, faam, all.x = TRUE)

faa.fit <- armatrix.fit(year=faam$year,age=faam$age,x=faam$fec,cv=faam$cv,shrink.cv = 0.5)  # shrink reduces cv variability, not overall uncertainty
faa.fit  # convergence ok

save(faa.fit,file=paste0("Rdata/fit.armatrix/",this.year,"base_cv1shrink0.5.Rdata"))
#save(faa.fit,file=paste0("Rdata/fit.armatrix/",this.year,"base_cv4shrink0.5.Rdata"))

```

## plots {.tabset}

### waa
```{r faa_final, message = F, fig.height = 3,fig.width = 5,results = 'hide',message=FALSE}
armatrix.waa(faa.fit,ylab='Fecundity')+scale_color_viridis_d()
```

### Predicted obs
```{r faa_predobs, message = F, fig.height = 10,fig.width = 10,results = 'hide',message=FALSE}
armatrix.predobs(faa.fit,scale='free',ncol=3)
```

### Predicted
```{r faa_pred, message = F, fig.height = 10,fig.width = 10,results = 'hide',message=FALSE}
armatrix.pred(faa.fit,scale='free',ncol=3)
```

### Resid bubble
```{r faa_res, message = F, fig.height = 3,fig.width = 10,results = 'hide',message=FALSE}
armatrix.res(faa.fit)
```

### Resid dot
```{r faa_res2, message = F, fig.height = 10,fig.width = 10,results = 'hide',message=FALSE}
armatrix.res2(faa.fit)
```

### effects
```{r faa_effects, message = F, fig.height = 10,fig.width = 10,results = 'hide',message=FALSE}
armatrix.effects(faa.fit)
```

### cvs
Only interesting if shrink argument is used.
```{r faa_cv, message = F, fig.height = 10,fig.width = 10,results = 'hide',message=FALSE}
armatrix.cvs(faa.fit,scale='free',ncol=3)
```

# Save results
Predict fecundity age 1 fish based on the predicted ratio of fecundity age 2 / age 1 from pelletier 
```{r faa_result, message = F}
faaf <- faa.fit$output
faaf <- faaf[,c(1:2,9)]
names(faaf)[3] <- 'fecundity'
faafm <- dcast(faaf,year~age,value.var = 'fecundity')

# add age 1
m <-  as.numeric(models[models$variable=='age',-1])
fec1fec2 <- (10^(m[1]+log10(1)*m[2]))/(10^(m[1]+log10(2)*m[2]))
fec1fec2

faafm <- cbind(year=faafm$year,'1'=faaf[faaf$age==2,'fecundity']*fec1fec2,faafm[,2:ncol(faafm)])
faafm[,2:ncol(faafm)] <- round(faafm[,2:ncol(faafm)],0)

write.csv(faafm, file=paste0('csv/faa_',this.year,'_base_cv1shrink0.5.csv'),row.names = FALSE)
#write.csv(faafm, file=paste0('csv/faa_',this.year,'_sens_cv4shrink0.5.csv'),row.names = FALSE)

p <- ggplot(melt(faafm,id='year'), aes(x = year, y = value/1000, col = variable))+ 
    geom_line()+
    labs(x = 'Year', y = 'Fecundity (thousands)', col = 'Age')+
    scale_x_continuous(expand = c(0, 0))+
    scale_color_viridis_d()
ggsave(filename = paste0('img/faa',this.year,'.png'),plot = p,units = 'cm',height = 8,width = 14)

```

